---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SAEforest

<!-- badges: start -->
<!-- badges: end -->

**This is a fork of [`krennpa/SAEforest`](https://github.com/krennpa/SAEforest)**
that adds:

* `var.adjust`: the variance bias correction of Krennmair et al. (2026, *Annals
  of Applied Statistics*), applied inside the MERF fitting loop.
* `transformation = "log"`: fit the MERF on `log(Y)` with smearing-based
  back-transformation of point/MSE estimates.
* `seed`: reproducibility for the whole `SAEforest_model()`/`MERFranger()`
  pipeline via a single top-level `set.seed()` call.
* `select.indicator`: compute only a subset of the standard nonlinear
  indicators, for faster estimation when just one or a few are needed.

`var.adjust` and `transformation` are scoped to the non-linear indicator path
that uses smearing (`meanOnly = FALSE`, `smearing = TRUE`); they are ignored
(with an explicit warning) under `meanOnly = TRUE`, `aggData = TRUE`, or
`smearing = FALSE`. See `vignette("variance-adjusted-merf")` for a worked
walkthrough of the new arguments. Upstream `krennpa/SAEforest` remains the
collaboration target for these changes via pull request.

**Known limitations (deferred, kept for upstream parity):** a verification of
`var.adjust` against Krennmair et al. (2026) identified two minor, low-priority
deviations from the paper that are intentionally left unchanged so this fork
stays a minimal diff against upstream: (P2) `adjust_ErrorSD()` builds its
bootstrap residuals from in-sample (non-OOB) forest predictions rather than
OOB predictions, biasing its bias term slightly low; (P3) the naive residual
variance in `MERFranger()` is computed as `sd(...)^2` (centred, denominator
`n - 1`) rather than the paper's uncentred `n^-1 Sum(.)^2`, which is
numerically negligible at typical SAE sample sizes but a literal departure
from the equation as written.

The package \pkg{SAEforest} promotes the use of Mixed Effects Random Forests (MERFs) for applications of Small Area Estimation (SAE). The package effectively combines functions for the estimation of regionally disaggregated linear and nonlinear economic and inequality indicators using survey sample data. Estimated models increase the precision of direct estimates from survey data, combining unit-level and aggregated population level covariate information from census or register data. Apart from point estimates, MSE estimates for requested indicators can be easily obtained. The package provides procedures to facilitate the analysis of model performance of MERFs and visualizes predictive relations from covariates and variable importance. Additionally, users can summarize and map indicators and corresponding measures of uncertainty.

## Installation

You can install the development version of SAEforest from Github with:

``` r
# install.packages("devtools")
devtools::install_github("krennpa/SAEforest")
```

## Example

This is a basic example which demonstrates the functionality of this package:

```{r example}
library(SAEforest)

data("eusilcA_pop")
data("eusilcA_smp")

income <- eusilcA_smp$eqIncome
X_covar <- eusilcA_smp[,-c(1,16,17,18)]

#Example 1:
#Calculating point estimates and discussing basic generic functions

model1 <- SAEforest_model(Y = income, X = X_covar, dName = "district",
                         smp_data = eusilcA_smp, pop_data = eusilcA_pop)

#SAEforest generics:
summary(model1)
```

I included some further features to inspect the model graphically. For instance look at the following output from the generic function `plot`, which shows a so-called variable importance plot:
```{r model1, echo = FALSE, out.width = "75%"}
plot(model1, pdp_plot = F)
```

We cannot only inspect the model graphically, but also map our indicators. Take a look at this example on Austrian pseudo-data for district-level mean income produced by the function `map_indicators`:
```{r , echo = FALSE, warning=FALSE, message=FALSE, results='hide', out.width = "75%"}
data("eusilcA_pop")
data("eusilcA_smp")

income <- eusilcA_smp$eqIncome
X_covar <- eusilcA_smp[,-c(1,16,17,18)]

#Example 1:
#Calculating point estimates and discussing basic generic functions

model1 <- SAEforest_model(Y = income, X = X_covar, dName = "district",
                       smp_data = eusilcA_smp, pop_data = eusilcA_pop, num.trees=200)

# Load shape file
load_shapeaustria()

# Create map plot for mean indicator - point and MSE estimates but no CV
map_indicators(object = model1, MSE = FALSE, CV = FALSE,
         map_obj = shape_austria_dis, indicator = c("Mean"),
         map_dom_id = "PB")
```

I hope you like this presentation and the package. If you are interested in model-based SAE you should definitely also check out package `emdi`.
