---
title: "Variance-Adjusted MERFs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance-Adjusted MERFs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This fork of `SAEforest` adds a small number of options to
`SAEforest_model()`/`MERFranger()` that are **not** present in the upstream
`krennpa/SAEforest` package:

* `var.adjust`: applies the variance bias correction of Krennmair et al.
  (2026, *Annals of Applied Statistics*) inside the MERF fitting loop.
* `transformation`: allows fitting the MERF on `log(Y)` and back-transforming
  point/MSE estimates via smearing.
* `seed`: seeds R's RNG once at the top of `SAEforest_model()` for
  reproducibility of the whole pipeline (point estimates and, if requested,
  the MSE bootstrap).
* `select.indicator`: restricts computation to a subset of the standard
  nonlinear indicators, which can meaningfully speed up estimation when only
  one or a few indicators (e.g. `"Hcr"`) are of interest.

**Scope limitation:** `var.adjust` and `transformation` only affect the
non-linear indicator path (`meanOnly = FALSE`) that uses the smearing
approach (`smearing = TRUE`, the default). Passing either argument together
with `meanOnly = TRUE`, `aggData = TRUE`, or `smearing = FALSE` now raises an
explicit `warning()` rather than silently ignoring the argument.

This vignette illustrates the new arguments on the bundled `eusilcA` example
data. Code chunks use a tiny `num.trees`/`B` and are not evaluated when the
vignette is built, so building the package never triggers a full,
computationally expensive MERF fit; copy the code into an interactive
session (with realistic `num.trees`/`B`) to reproduce it for real use.

```{r setup}
library(SAEforest)

data("eusilcA_pop")
data("eusilcA_smp")

income <- eusilcA_smp$eqIncome
X_covar <- eusilcA_smp[, -c(1, 16, 17, 18)]
```

## 1. A plain (unadjusted) fit

`var.adjust = FALSE` (the default) reproduces the original, unadjusted MERF
algorithm exactly.

```{r plain-fit, eval = FALSE}
model_plain <- SAEforest_model(
  Y = income, X = X_covar, dName = "district",
  smp_data = eusilcA_smp, pop_data = eusilcA_pop,
  meanOnly = FALSE, MSE = "none",
  var.adjust = FALSE,
  num.trees = 50, mtry = 3, seed = 1
)

summary(model_plain)
```

## 2. The variance-adjusted fit

Setting `var.adjust = TRUE` applies a per-iteration bias correction to the
random-effects residuals before refitting (Krennmair et al., 2026). This is
only meaningful on the non-linear, smearing-based path.

```{r var-adjust-fit, eval = FALSE}
model_adjusted <- SAEforest_model(
  Y = income, X = X_covar, dName = "district",
  smp_data = eusilcA_smp, pop_data = eusilcA_pop,
  meanOnly = FALSE, MSE = "none",
  var.adjust = TRUE, B_adj = 20,
  num.trees = 50, mtry = 3, seed = 1
)

summary(model_adjusted)
```

The bias correction estimates a bootstrap variance term `K` from `B_adj`
replications and subtracts it from the residual variance. With a very small
`B_adj` (as used here to keep the example fast), `K` can exceed the residual
variance; the adjusted standard deviation is then clamped to `0` and a warning
is emitted. Use an adequately large `B_adj` (e.g. 100 or more) in real
applications so the correction is stable.

## 3. A log-transformed fit

`transformation = "log"` fits the MERF on `log(Y)` and back-transforms via
smearing (non-finite back-transformed values become `NA`). It requires
strictly positive `Y`.

```{r log-fit, eval = FALSE}
model_log <- SAEforest_model(
  Y = income, X = X_covar, dName = "district",
  smp_data = eusilcA_smp, pop_data = eusilcA_pop,
  meanOnly = FALSE, MSE = "none",
  transformation = "log",
  num.trees = 50, mtry = 3, seed = 1
)

summary(model_log)
```

## Unsupported combinations

`var.adjust`/`transformation` are ignored (with a warning) outside the
non-linear smearing path, e.g.:

```{r unsupported-combo, eval = FALSE}
# Warns: var.adjust is ignored when meanOnly = TRUE
SAEforest_model(
  Y = income, X = X_covar, dName = "district",
  smp_data = eusilcA_smp, pop_data = eusilcA_pop,
  meanOnly = TRUE, var.adjust = TRUE,
  num.trees = 50, seed = 1
)
```

## References

Krennmair, P., & Schmid, T. (2022). Flexible Domain Prediction Using Mixed
Effects Random Forests. *Journal of the Royal Statistical Society: Series C
(Applied Statistics)*.

Krennmair, P., et al. (2026). Variance estimation for Mixed Effects Random
Forests in Small Area Estimation. *Annals of Applied Statistics* (forthcoming).

Mendez, G., & Lohr, S. (2011). Estimating residual variance in random forest
regression. *Computational Statistics & Data Analysis*, 55(11), 2937-2950.
